---
title: "week 12" 
format: 
  html: 
    embed-resources: true 
editor: visual
---

```{r}
#| message: FALSE
#| warning: false
#| cashe: false
#| echo: false

library(tidyverse)
library(gssr)
library(dplyr)
```

# Firstkid

```{r}
gss2018 <- gss_get_yr(2018)
firstkid <- gss2018 |>
  select(agekdbrn, educ, sex) |>
  mutate(latebirth = if_else(agekdbrn >= 28 , 1, 0),
         degree = if_else(educ >= 16, 1, 0),
         gender = if_else(sex == 1 , 1, 0)) |>
  drop_na()
```

```{r}
table <- firstkid |>
  group_by(degree, gender) |>
  summarize(
    n=n(), 
    p_latebirth = mean(latebirth),
    log_odds = log(p_latebirth/(1-p_latebirth))
  )
table
```

## Linear Regression

### Saturated Model

#### by hand

$$
P(latebirth_i) = \alpha + \beta_1 * degree_i + \beta_2 * gender_i + \beta_3 * degree_i * gender_i
$$

$\alpha$ = 0.137

For people having no college degree and being female, the probability of latebirth is 13.7%.

$\beta_1$ = 0.456 - 0.137 = 0.319

Holing gender constant, having college degree increases the probability of latebirth by 31.9 percentage points.

$\beta_2$ = 0.235 - 0.137 = 0.098

Holing degree constant, being male increases the probability of latebirth by 9.8 percentage points.

$\beta_3$ = 0.570 - 0.137 - 0.319 - 0.098 = 0.016

For male, having college degree increases the probability of latebirth by 1.6 percentage points; For college degree holders, being male increases the probability of latebirth by 1.6 percentage points. \#### by computer

```{r}
lm1 <- glm(latebirth ~ degree*gender,
           data = firstkid, 
           family = binomial(link = "identity"))
summary(lm1)
```

$$
P(latebirth_i) = 0.137 + 0.319 * degree_i + 0.098 * gender_i + 0.016 * degree_i * gender_i
$$

### Restricted Model

```{r}
lm2 <- glm(latebirth ~ degree+gender,
           data = firstkid, 
           family = binomial(link = "identity"))
summary(lm2)
```

$$
P(latebirth_i) = 0.136 + 0.326 * degree_i + 0.101 * gender_i
$$ 

For people without a college degree and being female, the probability of latebirth is 13.6%.

Holding gender constant, having college degree increases the probability of latebirth by 32.6 percentage points.

Holding degree constant, being male increases probability of latebirth by 10.1 percentage points.

### Compare

```{r}
anova(lm1, lm2)
```

The deviance of model with interaction term(1753.7) is smaller than model without(1753.8). The difference of two model is small (0.098). P value > 0.05, adding an interaction term doesn't provide a more fitted model. So restricted model is preferred.

## Logits Regression

### Saturate Model

#### by hand

$$
log(\frac{P(latebirth_i)}{1-P(latebirth_i)}) = \alpha + \beta_1*degree_i + \beta_2 *gender_i + \beta_3 * degree_i * gender_i
$$

$\alpha$ = -1.84

The log odds of latebirth of people who has no college degree(degree = 0) and is female(gender = 0) is -1.84

$\beta_1$ = -0.175 - (-1.84) = 1.665

Having college degree(degree = 1) increases log-odds of latebirth by 1.665

$\beta_2$ = -1.18 - (-1.84) = 0.66

Being male(gender = 1) increases log-odss of latebirth by 0.66

$\beta_3$ = 0.282 - (-1.84) - 1.665 - 0.66 = -0.203

For male, having college degree (degree = 1) decreases log-odds of latebirth by 0.203; For people with college degree, being male decreases log-odds of latebirth by 0.203.

#### by computer

```{r}
LR1 <- glm(latebirth ~ degree*gender,
          data = firstkid,
          family = binomial(link = "logit"))
summary(LR1)
```

$$
log(\frac{P(latebirth_i)}{1-P(latebirth_i)}) = -1.84 + 1.66 *degree_i + 0.66 *gender_i -0.20 * degree_i * gender_i
$$

### Restricted Model

```{r}
LR2 <- glm(latebirth ~ degree+gender,
          data = firstkid,
          family = binomial(link = "logit"))
summary(LR2)
```

$$
log(\frac{P(latebirth_i)}{1-P(latebirth_i)}) = -1.80 + 1.57 *degree_i + 0.57 *gender_i
$$

```{r}
exp(-1.80)
exp(1.57)
exp(0.57)
```

The odds ratio of latebirth of a person who has no college degree and is male is 0.158.

Having college degree increases odds ratio of having latebirth by 4.8 times.

Being a male increases odds ratio of having latebirth by 1.77 times.

### Compare

```{r}
anova(LR1, LR2)
```

The deviance of saturated model(1753.7) is smaller than restricted model(1754.4). But the difference is small(-0.71). P value > 0,05,  adding the interaction term doesn't make the model better. Usually the model the simpler, the better. So the restricted model is preferred.

## Poissons Regression

```{r}
tablep <- firstkid |>
  group_by(degree, gender) |>
  summarize(
    n=n(), 
    birthage = mean(agekdbrn),
  )
tablep
```

### Saturate Model

#### by hand

$$
birthage_i = \alpha + \beta_1*degree_i + \beta_2 *gender_i + \beta_3 * degree_i * gender_i
$$

$\alpha$ = 22.1

The birthage for people having no college degree and being female is 22.1 years old.

$beta_1$ = 26.8- 22.1 = 4.7

Having college degree increases birthage by 4.7 years.

$beta_2$ = 24.2 - 22.1 = 2.1

Being male increase birth age by 2.1 years.

$beta_3$ = 28.5 - 22.1 - 4.7 - 2.1 = -0.4

Having college degree and being male decreases birthage by 0.4 years.

#### by computer

```{r}
plm <- glm(agekdbrn ~ degree*gender,
           data = firstkid,
           family = poisson(link = "identity"))
summary(plm)
```

$$
birthage_i = 22.1 + 4.7*degree_i + 2.1 *gender_i - 0.3 * degree_i * gender_i
$$

#### Restricted Model

```{r}
plmr <- glm(agekdbrn ~ degree+gender,
           data = firstkid,
           family = poisson(link = "identity"))
summary(plmr)
```

$$
birthage_i = 22.11 + 4.55*degree_i + 2.02 *gender_i
$$

For people having no college degree and being female, age for first kid is 22.11 years old.

Holding gender constant, having college degree increases age for first kid by 4.55 years.

Holding degree constant, being male increases age for first kid by 2.02 years.

### Compare

```{r}
anova(plm, plmr)
```

Deviance of model with interaction term(1799.9) is smaller than model without(1800.2). But the difference is small. P value > 0.05, adding interaction term doesn't make the model better. Thus, restricted model is preferred.

# Hourswork

```{r}
hourswork <- gss2018 |>
  select(hrs1, educ, sex) |>
  mutate(overwork = if_else(hrs1 > 40 , 1, 0),
         degree = if_else(educ >= 16, 1, 0),
         gender = if_else(sex == 1 , 1, 0)) |>
  drop_na()
```

```{r}
table2 <- hourswork |>
  group_by(degree, gender) |>
  summarize(
    n=n(), 
    p_overwork = mean(overwork),
    log_odds = log(p_overwork/(1-p_overwork))
  )
table2
```

## Linear Regression

### Saturated Model

#### by hand

$$
P(overwork_i) = \alpha + \beta_1 * degree_i + \beta_2 * gender_i + \beta_3 * degree_i * gender_i
$$

$\alpha$ = 0.275

For people having no college degree and being female, the probability of work overtime is 27.5%.

$\beta_1$ = 0.322 - 0.275 = 0.047

Holing gender constant, having college degree increases the probability of work overtime by 4.7 percentage points.

$\beta_2$ = 0.497 - 0.275 = 0.222

Holing degree constant, being male increases the probability of work overtime by 22.2 percentage points.

$\beta_3$ = 0.504 - 0.275 - 0.047 - 0.222 = -0.04

For male, having college degree decreases the probability of work overtime by 4 percentage points; For college degree holders, being male decreases the probability of work overtime by 4 percentage points.

#### by computer

```{r}
llm1 <- glm(overwork ~ degree * gender,
            data = hourswork,
            family = binomial(link = "identity"))
summary(llm1)
```

$$
P(overwork_i) = 0.275 + 0.047 * degree_i + 0.222 * gender_i - 0.040 * degree_i * gender_i
$$

### Restricted Model

```{r}
llm2 <- glm(overwork ~ degree + gender,
            data = hourswork,
            family = binomial(link = "identity"))
summary(llm2)
```

$$
P(overwork_i) = 0.281 + 0.030 * degree_i + 0.207 * gender_i
$$

$\alpha$ = 0.281

For people having no college degree and being female, the probability of work overtime is 28.1%.

$\beta_1$ = 0.030

Holing gender constant, having college degree increases the probability of work overtime by 3 percentage points.

$\beta_2$ = 0.497 - 0.275 = 0.207

Holing degree constant, being male increases the probability of work overtime by 20.7 percentage points.

### Compare

```{r}
anova(llm1, llm2)
```

The deviance of saturated model is smaller than restricted model but the difference is small. In this case, adding an interaction term doesn't provide more accuracy. The simpler model (restricted) is preferred.

## Logits Regression

### Saturate Model

#### by hand

$$
log(\frac{P(overwork_i)}{1-P(overwork_i)}) = \alpha + \beta_1*degree_i + \beta_2 *gender_i + \beta_3 * degree_i * gender_i
$$

$\alpha$ = -0.969

The log odds of work overtime of people who has no college degree(degree = 0) and is female(gender = 0) is -0.969

$\beta_1$ = -0.744 - (-0.969) = 0.225

Having college degree(degree = 1) increases on log-odds of work overtime by 1.665

$\beta_2$ = -0.0138 - (-0.969) = 0.9552

Being male(gender = 1) increases log-odds of work overtime by 0.9552

$\beta_3$ = 0.0171 - (-0.969) - 0.225 - 0.95222 = -0.191

For male, having college degree (degree = 1) decreases log-odds of work overtime by 0.191; For college degree holders, being male decreases log-odds of work overtime by 0.191.

#### by computer
```{r}
LRR1 <- glm(overwork ~ degree*gender,
          data = hourswork,
          family = binomial(link = "logit"))
summary(LRR1)
```

$$
log(\frac{P(overwork_i)}{1-P(overwork_i)}) = -0.9694 + 0.2258*degree_i + 0.9556 *gender_i - 0.1949 * degree_i * gender_i
$$

### Restricted Model
```{r}
LRR2 <- glm(overwork ~ degree+gender,
          data = hourswork,
          family = binomial(link = "logit"))
summary(LRR2)
```

$$
log(\frac{P(overwork_i)}{1-P(overwork_i)}) = -0.9292 + 0.1245*degree_i + 0.8826 *gender_i
$$

```{r}
exp(-0.9292)
exp(0.1245)
exp(0.8826)
```

For people having no college degree and is female, the odds ratio of work overtime is 0.394.

Having college degree increases odds ratio by 1.13 times.

Being male increases odds ratio by 2.42 times.

### Compare
```{r}
anova(LRR1, LRR2)
```
The deviance of saturated model is smaller than restricted model but the difference is small. In this case, adding an interaction term doesn't provide more accuracy. The simpler model (restricted) is preferred.

## Poissons Regression

```{r}
tablep2 <- hourswork |>
  group_by(degree, gender) |>
  summarize(
    n=n(), 
    workhour = mean(hrs1),
  )
tablep2
```

### Saturate Model

#### by hand

$$
workhour_i = \alpha + \beta_1*degree_i + \beta_2 *gender_i + \beta_3 * degree_i * gender_i
$$

$\alpha$ = 37.9

The weekly work hour for people having no college degree and being female is 37.9 hours.

$beta_1$ = 38.7 - 37.9 = 0.8

Having college degree increases weekly work hour by 0.8 hours.

$beta_2$ = 44.6 - 37.9 = 6.7

Being male increase weekly work hour by 6.7 hours.

$beta_3$ = 44.7 - 37.9 - 0.8 - 6.7 = -0.7

For male, Having college degree decreases weekly work hour by 0.7 hours.
For college degree holders, being male decrease weekly work hour by 0.7 hours.

#### by computer

```{r}
plmm1 <- glm(hrs1 ~ degree*gender,
           data = hourswork,
           family = poisson(link = "identity"))
summary(plmm1)
```

$$
workhour_i = 37.85 + 0.81*degree_i + 6.79 *gender_i - 0.79 * degree_i * gender_i
$$
### Restricted
```{r}
plmm2 <- glm(hrs1 ~ degree+gender,
           data = hourswork,
           family = poisson(link = "identity"))
summary(plmm2)
```

$$
workhour_i = 37.98 + 0.47*degree_i + 6.50 *gender_i
$$

For people having no college degree and is female, the weekly work hour is 37.98 hours.

Having college degree increases weekly work hour by 0.47 hours.

Being male increases weekly work hour by 6.5 hours.

### Compare
```{r}
anova(plmm1, plmm2)
```
The deviance of saturated model is smaller than restricted model but the difference is small. In this case, p value > 0.05, adding an interaction term doesn't provide more accuracy. The simpler model (restricted) is preferred. 

